Zero-shot task adaptation by homoiconic meta-mapping

### 2019-11-07 writing/homoiconic_meta_mapping_main.tex
Adding some things suggested by reviewers

\vspace{-0.75em} % section
\section{Introduction}
\vspace{-0.5em} % section
Humans are able to use and reuse knowledge more flexibly than most deep learning models can \citep{Lake2016, Marcus2018}. The problem of rapid learning has been partially addressed by meta-learning systems \citep[see also section \ref{sec_discussion}]{Santoro2016, Finn2017a, Finn2018, Stadie2018, Botvinick2019}. However, humans can use our knowledge of a task to flexibly adapt when the task changes. In particular, we can often perform an altered task zero-shot, that is, without seeing any data at all. For example, once we learn to play a game, we can immediately switch to playing in order to lose, and can perform reasonably on our first attempt. \par
One fundamental reason for this is that humans are aware of what we are trying to compute and why. This allows us to adapt our task representations to perform a new task zero-shot. By contrast, most deep learning models do not explicitly represent tasks at all. Those that do (such as some meta-learning models) generally do not represent relationships between tasks, and so cannot adapt to a new task without data from the new task. To address this, it is necessary to represent tasks in a transformable space. This can grant the ability to rapidly adapt behavior to a new task. \par
In this paper, we propose a new class of tasks based on this idea: meta-mappings, i.e. mappings between tasks (see below). As noted above, this type of transfer is easily accessible to humans \citep{Lake2016}, but is generally inaccessible to deep-learning models. To address this challenge, we propose using architectures which essentially take a functional perspective on meta-learning, and exploit the idea of homoiconicity. (A homoiconic programming language is one in which programs in the language can be manipulated by programs in the language, just as data can.) By treating both data and task behaviors as functions, we can conceptually think of both data \emph{and} learned task behaviors as transformable. This yields the ability to not only learn to solve new tasks, but to learn how to transform these solutions in response to changing task demands. We demonstrate that our architectures can flexibly remap their behavior to address the meta-mapping challenge. By allowing the network to recursively treat its task representations as data points, and transform them to produce new task representations, our approach is able to achieve this flexibility parsimoniously. We suggest that approaches like ours will be key to building more intelligent and flexible deep learning systems. \par

@@

\paragraph{Related work:} Our work is an extrapolation from the rapidly-growing literature on meta-learning \citep[e.g.][]{Vinyals2016, Santoro2016, Finn2017a, Finn2018, Stadie2018, Botvinick2019}. It is also related to the literature on continual learning, or more generally tools for avoiding catastrophic interference based on changes to the architecture \citep[e.g.][]{Fernando2017, Rusu2016}, loss \citep[e.g.][]{Kirkpatrick2016, Zenke2017, Aljundi2019}, or external memory \citep[e.g.][]{Sprechmann2018}. We also connect to a different perspective on continual learning in appendix \ref{app_continual}. Recent work has also begun to blur the separation between these approaches, for example by meta-learning in an online setting \citep{Finn2019}. Our work is specifically inspired by the algorithms that attempt to have the system learn to adapt to a new task via activations rather than weight updates, either from examples \citep[e.g.][]{Wang2016a, Duan2016}, or a task input \citep[e.g.][]{Borsa2019}. \par
Our architecture builds directly off of prior work on HyperNetworks \citep{Ha2016} -- networks which parameterize other networks -- and other recent applications thereof, such as guessing parameters for a model to accelerate model search \citep[e.g.][]{Brock2018a, Zhang2019}, and meta-learning \citep[e.g.]{Li2019a, Rusu2019}. Our work is also related to the longer history of work on different time-scales of weight adaptation \citep{Hinton1982, Kumaran2016} that has more recently been applied to meta-learning contexts \citep[e.g.][]{Ba2016, Munkhdalai2017, Garnelo2018} and continual learning \citep[e.g.]{Hu2019}. It is more abstractly related to work on learning to propose architectures \citep[e.g.][]{Zoph2016, Cao2019}, and to models that learn to select and compose skills to apply to new tasks \citep[e.g.][]{Andreas, Andreas2016, Tessler2016, Reed2015, Chang2019a}. In particular, some of the work in domains like visual question answering has explicitly explored the idea of building a classifier conditioned on a question \citep{Andreas, Andreasa}, which is related to one of the possible computational paths through our architecture. Work in model-based reinforcement learning has also partly addressed how to transfer knowledge between different reward functions \citep[e.g.][]{Laroche2017}; our approach is more general. Indeed, our insights could be combined with model-based approaches, for example our approach could be used to adapt a task embedding, which would then be used by a learned planning model.\par 
There has also been other recent interest in task (or function) embeddings. Achille et al. \citep{Achille2019} recently proposed computing embeddings for visual tasks from the Fisher information of the parameters in a model partly tuned on the task. They show that this captures some interesting properties of the tasks, including some types of semantic relationships, and can help identify models that can perform well on a task. Rusu and colleagues recently suggested a similar meta-learning framework where latent codes are computed for a task which can be decoded to a distribution over parameters \citep{Rusu2019}. Other recent work has tried to learn representations for skills \citep[e.g.][]{Eysenbach2019} or tasks \citep[e.g.]{Hsu2019} for exploration and representation learning. Our perspective can be seen as a generalization of these that allows for remapping of behavior via meta-tasks. To the best of our knowledge none of the prior work has explored zero-shot performance of a task via meta-mappings. \par

@@

We've highlighted a new type of flexibility in the form of meta-mapping between tasks. Meta-mapping can produce zero-shot performance on novel tasks, based on their relationship to old tasks. This is a key aspect of human flexibility, but is lacking from most contemporary deep-learning models. Achieving this flexibility requires representing tasks in a transformable way. To address this, our Homoiconic Meta-Mapping (HoMM) approach explicitly represents tasks by function embeddings, and derives the computation from these embeddings via a HyperNetwork. This allows for our key insight: by embedding tasks and individual data points in a shared latent space, the same meta-learning architecture can be used both to transform data for basic tasks, and to transform task embeddings to adapt to task variations. This approach is parsiomious, because it uses the same networks for basic- and meta-mappings. Perhaps because of this, the requisite meta-task sample complexity is small; we showed generalization to unseen meta-mappings after training on only 20 meta-mappings. That is, we are able to achieve zero-shot performance on a novel task based on a held-out relationship between tasks. This is a step closer to human-level flexibility.\par

>>>>

\vspace{-0.75em} % section
\section{Introduction}
\vspace{-0.5em} % section
Humans are able to use and reuse knowledge more flexibly than most deep learning models can \citep{Lake2016, Marcus2018}. The problem of rapid learning has been partially addressed by meta-learning systems \citep[see also section \ref{sec_discussion}]{Santoro2016, Finn2017a, Finn2018, Stadie2018, Botvinick2019}. However, humans can use our knowledge of a task to flexibly adapt when the task changes. In particular, we can often perform an altered task zero-shot, that is, without seeing any data at all. For example, once we learn to play a game, we can immediately switch to playing in order to lose, and can perform reasonably on our first attempt. \par
One fundamental reason for this is that humans are aware of what we are trying to compute and why. This allows us to adapt our task representations to perform an altered task. By contrast, most prior work on zero-shot learning relies on generating a completely new representation for a new task, e.g. from a natural language description \citep[e.g.]{Socher2013, Oh2017a}. These systems must therefore throw away much of their prior knowledge of the task and regenerate it from scratch, rather than just transforming what needs to be transformed. Other approaches use relationships between tasks as an auxiliary learning signal \citep[e.g.]{Yao2019, Pal2019}, but cannot adapt to a new task by transforming their task representations. To address this, it is necessary to represent tasks in a transformable space. This can grant the ability to rapidly and precisely adapt behavior to a new task. \par
In this paper, we propose a new class of tasks based on this idea: meta-mappings, i.e. mappings between tasks (see below). Meta-mappings allow zero-shot performance of a new task based on its relationship to old tasks. To address the challenge of performing meta-mappings, we propose using architectures which essentially take a functional perspective on meta-learning, and exploit the idea of homoiconicity. (A homoiconic programming language is one in which programs in the language can be manipulated by programs in the language, just as data can.) By treating both data and task behaviors as functions, we can conceptually think of both data \emph{and} learned task behaviors as transformable. This yields the ability to not only learn to solve new tasks, but to learn how to transform these solutions in response to changing task demands. We demonstrate that our architectures can flexibly remap their behavior to address the meta-mapping challenge. By allowing the network to recursively treat its task representations as data points, and transform them to produce new task representations, our approach is able to achieve this flexibility parsimoniously. We suggest that approaches like ours will be key to building more intelligent and flexible deep learning systems. \par

@@

\paragraph{Related work:} There has been a variety of past work on zero-shot learning, mostly based on natural-language descriptions of tasks. For example, \citet{Larochelle2008} considered the general problem of behaving in accordance with language instructions as simply asking a model to adapt its response when conditioned on different ``instruction'' inputs. Later work explored zero-shot classification based on only a natural language description of the target class \citep{Socher2013,Romera-Paredes2015,Xian2018}, or of a novel task in a language-conditioned RL system \citep{Hermann2017}. Some work has even exploited relationships between tasks as a learning signal \citep{Oh2017a, Pal2019}. However, to our knowledge none of the prior work has proposed using meta-mapping approaches to adapt to new tasks. \par
Our work is an extrapolation from the rapidly-growing literature on meta-learning \citep[e.g.][]{Vinyals2016, Santoro2016, Finn2017a, Finn2018, Stadie2018, Botvinick2019}. It is also related to the literature on continual learning, or more generally tools for avoiding catastrophic interference based on changes to the architecture \citep[e.g.][]{Fernando2017, Rusu2016}, loss \citep[e.g.][]{Kirkpatrick2016, Zenke2017, Aljundi2019}, or external memory \citep[e.g.][]{Sprechmann2018}. We also connect to a different perspective on continual learning in appendix \ref{app_continual}. Recent work has also begun to blur the separation between these approaches, for example by meta-learning in an online setting \citep{Finn2019}. Our work is specifically inspired by the algorithms that attempt to have the system learn to adapt to a new task via activations rather than weight updates, either from examples \citep[e.g.][]{Wang2016a, Duan2016}, or a task input \citep[e.g.][]{Borsa2019}. \par
Our architecture builds directly off of prior work on HyperNetworks \citep{Ha2016} -- networks which parameterize other networks -- and other recent applications thereof, such as guessing parameters for a model to accelerate model search \citep[e.g.][]{Brock2018a, Zhang2019}, and meta-learning \citep[e.g.]{Li2019a, Rusu2019}. In particular, recent work in natural language processing has shown that having contextually generated parameters can allow for zero-shot task performance, assuming that a good representation for the novel task is given \citep{Platanios2017} -- in their work this representation was evident from the factorial structure of translating between many pairs of languages. Our work is also related to the longer history of work on different time-scales of weight adaptation \citep{Hinton1982, Kumaran2016} that has more recently been applied to meta-learning contexts \citep[e.g.][]{Ba2016, Munkhdalai2017, Garnelo2018} and continual learning \citep[e.g.]{Hu2019}. It is more abstractly related to work on learning to propose architectures \citep[e.g.][]{Zoph2016, Cao2019}, and to models that learn to select and compose skills to apply to new tasks \citep[e.g.][]{Andreas, Andreas2016, Tessler2016, Reed2015, Chang2019a}. In particular, some of the work in domains like visual question answering has explicitly explored the idea of building a classifier conditioned on a question \citep{Andreas, Andreasa}, which is related to one of the possible computational paths through our architecture. Work in model-based reinforcement learning has also partly addressed how to transfer knowledge between different reward functions \citep[e.g.][]{Laroche2017}; our approach is more general. Indeed, our insights could be combined with model-based approaches, for example our approach could be used to adapt a task embedding, which would then be used by a learned planning model.\par 
There has also been other recent interest in task (or function) embeddings. Achille et al. \citep{Achille2019} recently proposed computing embeddings for visual tasks from the Fisher information of the parameters in a model partly tuned on the task. They show that this captures some interesting properties of the tasks, including some types of semantic relationships, and can help identify models that can perform well on a task. Rusu and colleagues recently suggested a similar meta-learning framework where latent codes are computed for a task which can be decoded to a distribution over parameters \citep{Rusu2019}. Other recent work has tried to learn representations for skills \citep[e.g.][]{Eysenbach2019} or tasks \citep[e.g.]{Hsu2019} for exploration and representation learning. Our perspective can be seen as a generalization of these that allows for remapping of behavior via meta-tasks. While there have been other approaches to zero-shot task performance, to the best of our knowledge none of the prior work has explored zero-shot performance of a task via meta-mappings. \par

@@

We've highlighted a new type of flexibility in the form of meta-mapping between tasks. Meta-mapping can produce zero-shot performance on novel tasks, based on their relationship to old tasks. This is a key aspect of human flexibility. The meta-mapping perspective suggests that this flexibility requires representing tasks in a transformable way. To address this, our Homoiconic Meta-Mapping (HoMM) approach explicitly represents tasks by function embeddings, and derives the computation from these embeddings via a HyperNetwork. This allows for our key insight: by embedding tasks and individual data points in a shared latent space, the same meta-learning architecture can be used both to transform data for basic tasks, and to transform task embeddings to adapt to task variations. This approach is parsiomious, because it uses the same networks for basic- and meta-mappings. Perhaps because of this, the requisite meta-task sample complexity is small; we showed generalization to unseen meta-mappings after training on only 20 meta-mappings. That is, we are able to achieve zero-shot performance on a novel task based on a held-out relationship between tasks. Furthermore, the meta-mapping approach yields better task performance than a zero-shot approach based on performing the task based on just a natural language description (supplementary fig. \ref{fig_why_meta_mapping}). This is a step closer to human-level flexibility. \par
